3D Regression Heat Map Analysis of Population Study Data

### 2015-06-23 draft.tex
shortened the paper

\paragraph{Logistic Regression for Dichotomous Target.} Logistic regression implies a dichotomous target variable.
%%
The target is described by fitting a logistic function.
%%
Logistic models, as opposed to linear models, do not allow for extracting an $R^2$ quality of fit value.
%%
Therefore, pseudo-$R^2$ values are extracted, such as the \emph{Nagelkerke $R^2$}, which mimics the behavior of the $R^2$.
%%
\emph{Nagelkerke $R^2$} behaves different than $R^2$ values extracted from the linear regression model.
%%
Comparisons have to be handled with care.
%%
\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
%It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per participant.
%%
A second cohort, \texttt{SHIP-TREND} was started in 2012.
%%
Data for both cohorts are examined in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the latest two acquisitions (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
However, most examinations occur in all stages and are performed according to the same instructions to enable comparisons over the different stages of the study.
%%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
In 1977, Tukey already stated that data are too often analyzed solely using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses instead, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
%The work of M\"{u}hlbacher et al.~\cite{Muehlbacher}, Piringer et al.~\cite{Piringer}, Chan et al.~\cite{Chan} and Zhang et al.~\cite{Zhang2014} are closest related to ours regarding different aspects.
\paragraph{Parameter space analysis using regression models.}
%%
\magenta{
Sedlmair et al.~\cite{Sedlmair} present a conceptual taxonomy of parameter space analysis.
%%
%It includes a abstract application domain independent data flow model, navigation strategies for exploring parameter spaces as well a characterization of analysis tasks.
%%
Using the taxonomy, approaches can be classified to find works with similar problems and derive new design ideas and evaluate ideas.
%The taxonomy allows to classify approaches to find works with similar problems to derive new design ideas and evaluate ideas.
%%
Based on their input parameter taxonomy, we are visualizing \emph{model parameters} based on \emph{environmental parameters} in a \emph{global-to-local} navigation strategy.
%%
We solve a \emph{fitting} task by aiming to find models well suited for describing the input data.
}

@@

The difference of the presented approach in comparison with our previous and related work is twofold:

@@

\begin{enumerate}

@@

	\item We focus on the large-scale analysis of a vast number of linear and logistic regression-models by assessing their quality-of-fit using descriptive metrics.

@@

	\item The analysis is conducted w.r.t. a target feature and incorporates expert knowledge via the regression model definition rather than subdividing the underlying data.
\end{enumerate}

@@

In a data set with $n$ (e.g., 100) features, the regression formula
\begin{equation}
Cancer \sim X + Y
\end{equation}

@@

%% INTERACTION
Transformation of the 3D heat map is restricted to the y-axis (horizontal only), preserving the mental map to position individual features.
%%
The 3D heat map is always oriented according to the 2D representation, allowing for an easy mental combination of them.
%%
Allowing more degrees of freedom was confusing to our users and also did not add value to the visualization.
%%
Also we provide a zoom functionality using the mouse wheel input.
%%
\paragraph{3D heat map slice selection [\textbf{3.ZF}].}
%%
In order to \emph{Zoom, Filter and Analyze Further}, the user has to navigate to different slices of interest.
%%
We propose two ways to achieve this.

@@

\section{Implementation} \label{implementation}
\begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/implementation}
 \caption{
 The front-end (left) is realized with \emph{HTML5}/\emph{CSS3}/\emph{Javascript} and different Javascript libraries, such as \emph{Angular.js}, \emph{Three.js} and D3.js.
 %%
 The web-server back-end (right) is written using \emph{Node.js}, hosting is carried out using \emph{Heroku}.
 %%
 \emph{R} and \emph{OpenCPU} constitute the statistical back-end (top) to compute the \emph{3D Regression Heat Maps}.
 %%
 Additional statistical back-ends can be attached to the system to decrease the computation time.
 %%
 }
  \label{fig:Implementation}
\end{figure}
%%
We rely on web-based technologies for our prototype.
%%
%The ongoing transition of open-science software into the web spawned numerous projects, making state-of-the-art algorithms available in this domain.
%%
%\paragraph{Front-End.}
%%
The front-end is created using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%%
\texttt{Angular.js}\footnote{Open Source; Maintained by Google, \href{https://www.angularjs.org/}{\texttt{angularjs.org}}} abstracts web application into models and views, allowing for a responsive way to combine \texttt{HTML} and \texttt{Javascript}.
%%
In is easily expandable by forcing developers to write modularized code.
%It forces developers to write modularized code, which makes the components easier expandable. %while keeping the code maintainable by including unit tests.
%%
The page layout is handled using \texttt{Twitter Bootstrap}\footnote{Open Source; Maintained by Twitter, \href{http://getbootstrap.com}{\texttt{getbootstrap.com}}}, which also provides a rich set of user interface elements. %with proper stylings.
%%
The 2D heat map is implemented using \texttt{Data driven Documents} (\texttt{D3.js}), which is popular in information visualization using vector graphics \cite{D3}.
%%
It provides fast and easy methods for binding data to graphical elements.
%%
The 3D plot is created using the WebGL-based \texttt{Threejs}\footnote{Open Source; Originally developed by R. Cabello, \href{http://threejs.org}{\texttt{threejs.org}}} library.
%%
% We experimented with different ways for achieving the cube representation, including volume rendering, cube primitives for each data point and shader-based solutions.
% %%
% Open source volume rendering methods are available but do not satisfy our requirements.
% %%
% Creating a cube primitive for each data point resulted in non-interactive frame-rates for data sets larger than 30 features (creating 30$^3$ cube primitives).
% %%
% Therefore, we decided to use a shader-based solution by rendering the cube as a sprite-based particle system, allowing to customize color and opacity of every data point.
% %%
% It also is the fastest solution that we tested.

%\paragraph{Back-End.}
Two server structures serve as back-end.
%%
The web-server is written in \texttt{Javascript} using \texttt{Node.js}\footnote{Open Source; Maintained by Joyent Inc, \href{http://nodejs.org}{\texttt{nodejs.org}}}, running on Googles V8 Javascript runtime environment.
%%
It is hosted on \texttt{Heroku}\footnote{Owned by Salesforce.com, \href{https://www.heroku.com/}{\texttt{heroku.com}}}, a cloud application platform.
%%

The statistical computations are performed on the second structure.
%%
They rely on the statistical programming language \texttt{R}.\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}}
%%
It is widely adopted in the statistical analysis community, yielding a rich support of fast state-of-the-art statistics algorithms as well newly published methods.
%%
\texttt{OpenCPU} is an \texttt{R} package and provides an API for accessing it via HTTP calls \cite{Ooms}.
%%
This way, any computer, which runs \texttt{R} can be turned into a statistical processor for our project.
%%
% The back-end functions necessary for all cube calculations are provided via an \texttt{R} package.
% %%
% It uses multi-core optimization to use all machine CPUs to speed up the calculation process.
% %%
% The server workload balances are managed by the front-end code.

@@

Both data sets are unusual for epidemiological analysis regarding their feature extent.
%%
Usually, only a few features depicting a hypothesis are compiled into a data set to assess them using statistical tools.
%%
The herein used data sets comprise several hundred features.
%%
Our method focuses on data exploration and knowledge extraction and requires a wide scope of sociodemographic, medical and lifestyle features.
%%
\subsection{Participants, Setup and Procedure}
%%
The ability of a system to discover knowledge is difficult to measure.
%%
Lam et al.~\cite{Lam2012} propose the \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}) technique, which is focused on the characterization of a systems' ability to generate hypotheses and explore the data in order to extract information.
%%
\emph{VDAR} can be carried out based on case studies using thinking-aloud techniques to comprehend the reasoning and thought process of the user.
%%
We employ \emph{VDAR} for analyzing our system.
%%
\paragraph{Participants, setup and procedure.}
%%
We conducted a web-based analysis by using an online-meeting software, which features voice chat as well as screen sharing.
%%
Starting an analysis using these techniques takes about 5-10 minutes of setup time.

The sessions started with an initial overview of the system, showcasing its features and functionality.
%%
Afterwards, the experts use the system on their own computers.
%%
The screen-sharing function was still used to observe the actions of the expert.
%%
All sessions were video recorded to be processed later on.
%%

%%
We conducted the analysis with three participants.
%%
\emph{KH}, a clinician (10 years of experience) with focus on epidemiological research, is the domain expert for the breast fat data set.
%%
She is a radiologist, responsible for the SHIP-MRI acquisition and also responsible of the mammography analysis.
%%
The hepatic steatosis data set is analyzed by \emph{UN}, a data scientist responsible for prior analysis of the data.
%%
The third participant is \emph{TI}, a statistician with focus on epidemiology (8 years of experience) assesses the statistical reliability of the tool and the underlying methods without a focus on a specific data set.
%%
\subsection{The Hepatic Steatosis Data Set}
We employ the data set used by Niemann et al.~\cite{Niemann2014} to identify predictive features w.r.t. the reversible hepatic steatosis disorder.
%%
The dichotomous target feature is derived from the liver fat concentration measured using MRI scans.
%%
Liver fat concentrations of no more than 10\% are mapped to the `negative' class; values greater than 10\% are mapped to the `positive' class to indicate absence or presence of the disease.
%%
The data set contains labels for 578 participants. %for which the MRI data was available to that point.
%%
The MRI scans for each subject are only available in SHIP-2. %, so the target feature is only present the last study wave.

@@

As a next step, we want to introduce more regression types, which model different kinds of correlations.
%%
We also want to extend the 3D heat map to time-dependent data by expanding the difference heat map approach.
%%

We published all associated code and provide a freely accessible analysis platform open to heterogenous data types.
%%
We want to support opening up knowledge discovery to allow a diverse group of domain experts to derive insight into their data. %and support the notion of open science.
%%
\begin{small}
\acknowledgments{
%%
SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania.
%%
Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern.
%%
%The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG.
%%
This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.
%%
Sylvia Gla{\ss}er and Kai Lawonn are funded by the Federal Ministry of Education and Research within the Forschungs-campus STIMULATE under grant number '13GW0095A'.
}
\end{small}
\clearpage
\newpage
\bibliographystyle{abbrv}
%%use following if all content of bibtex file should be shown
%\nocite{*}
\bibliography{bibliography}
\end{document}

>>>>

\paragraph{Logistic Regression for Dichotomous Target.} Logistic regression implies a dichotomous target variable.
%%
The target is described by fitting a logistic function.
%%
Logistic models, as opposed to linear models, do not allow for extracting an $R^2$ quality of fit value.
%%
Therefore, pseudo-$R^2$ values are extracted, such as the \emph{Nagelkerke $R^2$}, which mimics the behavior of the $R^2$.
%%
\emph{Nagelkerke $R^2$} behaves different than $R^2$ values extracted from the linear regression model.
%%
Comparisons have to be handled with care.
%%
\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
%It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per participant.
%%
A second cohort, \texttt{SHIP-TREND} was started in 2012.
%%
Data for both cohorts are examined in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the latest two acquisitions (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
However, most examinations occur in all stages and are performed according to the same instructions to enable comparisons over the different stages of the study.
%%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
Tukey already stated in 1977 that data are too often analyzed solely using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to \emph{derive} hypotheses, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
%The work of M\"{u}hlbacher et al.~\cite{Muehlbacher}, Piringer et al.~\cite{Piringer}, Chan et al.~\cite{Chan} and Zhang et al.~\cite{Zhang2014} are closest related to ours regarding different aspects.
\paragraph{Parameter space analysis using regression models.}
%%
\magenta{
Sedlmair et al.~\cite{Sedlmair} present a conceptual taxonomy of parameter space analysis.
%%
%It includes a abstract application domain independent data flow model, navigation strategies for exploring parameter spaces as well a characterization of analysis tasks.
%%
Using the taxonomy, approaches can be classified to find works with similar problems and derive new design ideas and evaluate ideas.
%The taxonomy allows to classify approaches to find works with similar problems to derive new design ideas and evaluate ideas.
%%
Based on their input parameter taxonomy, we are visualizing \emph{model parameters} based on \emph{environmental parameters} in a \emph{global-to-local} navigation strategy.
%%
We solve a \emph{fitting} task by aiming to find models well suited for describing the input data.
}

@@

The difference of the presented approach in comparison with our previous and related work is twofold.

@@

(1) We focus on the large-scale analysis of a vast number of linear and logistic regression-models by assessing their quality-of-fit using descriptive metrics.

@@

(2) The analysis is conducted w.r.t. a target feature and incorporates expert knowledge via the regression model definition rather than subdividing the underlying data.

@@

% \begin{enumerate}
% 	\item We focus on the large-scale analysis of a vast number of linear and logistic regression-models by assessing their quality-of-fit using descriptive metrics.
% 	\item The analysis is conducted w.r.t. a target feature and incorporates expert knowledge via the regression model definition rather than subdividing the underlying data.
% \end{enumerate}

@@

In a data set with $n$ (e.g., 100) features, the regression formula $Cancer \sim X + Y$
%\begin{equation}
%
%\end{equation}

@@

%% INTERACTION
Transformation of the 3D heat map is restricted to the y-axis (horizontal only), preserving the mental map to position individual features.
%%
The 3D heat map is always oriented according to the 2D representation, allowing for an easy mental combination of them.
%%
Allowing more degrees of freedom was confusing to our users and also did not add value to the visualization.
%%
%Also we provide a zoom functionality using the mouse wheel input.
%%
\paragraph{3D heat map slice selection [\textbf{3.ZF}].}
%%
In order to \emph{Zoom, Filter and Analyze Further}, the user has to navigate to different slices of interest.
%%
We propose two ways to achieve this.

@@

\section{Implementation} \label{implementation}
\begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/implementation}
 \caption{
 The front-end (left) is realized with \emph{HTML5}/\emph{CSS3}/\emph{Javascript} and different Javascript libraries, such as \emph{Angular.js}, \emph{Three.js} and D3.js.
 %%
 The web-server (right) is written using \emph{Node.js} and hosted on \emph{Heroku}.
 %%
 \emph{R} and \emph{OpenCPU} constitute the statistical back-end (top) to compute the \emph{3D Regression Heat Maps}.
 %%
 Additional statistical back-ends can be attached to the system to decrease the computation time.
 %%
 }
  \label{fig:Implementation}
\end{figure}
%%
We rely on web-based technologies for our prototype.
%%
%The ongoing transition of open-science software into the web spawned numerous projects, making state-of-the-art algorithms available in this domain.
%%
%\paragraph{Front-End.}
%%
The front-end is created using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%%
%\texttt{Angular.js}\footnote{Open Source; Maintained by Google, \href{https://www.angularjs.org/}{\texttt{angularjs.org}}} abstracts web application into models and views, allowing for a responsive way to combine \texttt{HTML} and \texttt{Javascript}.
\texttt{Angular.js} abstracts web application into models and views, allowing for a responsive way to combine \texttt{HTML} and \texttt{Javascript}.
%%
In is easily expandable by forcing developers to write modularized code.
%It forces developers to write modularized code, which makes the components easier expandable. %while keeping the code maintainable by including unit tests.
%%
%The page layout is handled using \texttt{Twitter Bootstrap}\footnote{Open Source; Maintained by Twitter, \href{http://getbootstrap.com}{\texttt{getbootstrap.com}}}, which also provides a rich set of user interface elements.
\texttt{Twitter Bootstrap} handles the page layout and provides a rich set of user interface elements.
%%
The 2D heat map is implemented using the \texttt{D3.js}~\cite{D3} information visualization library.
%%
It provides fast and easy methods for binding data to graphical elements.
%%
%The 3D plot is created using the WebGL-based \texttt{Threejs}\footnote{Open Source; Originally developed by R. Cabello, \href{http://threejs.org}{\texttt{threejs.org}}} library.
The 3D plot is created using the WebGL-based library.
%%
% We experimented with different ways for achieving the cube representation, including volume rendering, cube primitives for each data point and shader-based solutions.
% %%
% Open source volume rendering methods are available but do not satisfy our requirements.
% %%
% Creating a cube primitive for each data point resulted in non-interactive frame-rates for data sets larger than 30 features (creating 30$^3$ cube primitives).
% %%
% Therefore, we decided to use a shader-based solution by rendering the cube as a sprite-based particle system, allowing to customize color and opacity of every data point.
% %%
% It also is the fastest solution that we tested.

%\paragraph{Back-End.}
Two server structures serve as back-end.
%%
%The web-server is written in \texttt{Javascript} using \texttt{Node.js}\footnote{Open Source; Maintained by Joyent Inc, \href{http://nodejs.org}{\texttt{nodejs.org}}}, running on Googles V8 Javascript runtime environment.
The web-server is written in \texttt{Javascript} using \texttt{Node.js}, running on Googles V8 Javascript runtime environment.
%%
%It is hosted on \texttt{Heroku}\footnote{Owned by Salesforce.com, \href{https://www.heroku.com/}{\texttt{heroku.com}}}, a cloud application platform.
It is hosted on \texttt{Heroku}, a cloud application platform.
%%

The statistical computations are performed on the second structure.
%%
%They rely on the statistical programming language \texttt{R}.\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}}
They rely on the statistical programming language \texttt{R}.
%%
It is widely adopted in the statistical analysis community, yielding a rich support of fast state-of-the-art statistics algorithms as well newly published methods.
%%
\texttt{OpenCPU} is an \texttt{R} package and provides an API for accessing it via HTTP calls \cite{Ooms}.
%%
This way, any computer, which runs \texttt{R} can be turned into a statistical processor for our project.
%%
% The back-end functions necessary for all cube calculations are provided via an \texttt{R} package.
% %%
% It uses multi-core optimization to use all machine CPUs to speed up the calculation process.
% %%
% The server workload balances are managed by the front-end code.

@@

Both data sets are unusual for epidemiological analysis regarding their feature extent.
%%
Usually, only a few features depicting a hypothesis are compiled into a data set to assess them using statistical tools.
%%
The herein used data sets comprise several hundred features.
%%
Our method focuses on data exploration and knowledge extraction and requires a wide scope of sociodemographic, medical and lifestyle features.
%%
\subsection{Participants, Setup and Procedure}
%%
The knowledge discovery capabilities of a system are difficult to measure.
%%
The \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}) technique proposed by Lam et al.~\cite{Lam2012} is focused on the characterization of a systems' ability to generate hypotheses and explore the data in order to extract information.
%%
\emph{VDAR} can be carried out based on case studies using thinking-aloud techniques to comprehend the users reasoning and thought process.
%%
We employ \emph{VDAR} for analyzing our system.
%%
\paragraph{Participants, setup and procedure.}
%%
We conducted a web-based analysis by using an online-meeting software, which features voice chat as well as screen sharing.
%%
Starting an analysis using these techniques takes about 5-10 minutes of setup time.
%%
The sessions started with an initial overview of the system, showcasing its features and functionality.
%%
Afterwards, the experts use the system on their own computers.
%%
The screen-sharing function was still used to observe the actions of the expert.
%%
All sessions were video recorded to be processed later on.
%%
We conducted the analysis with three participants.
%%
\emph{KH}, a clinician (10 years of experience) with focus on epidemiological research, is the domain expert for the breast fat data set.
%%
She is a radiologist, responsible for the SHIP-MRI acquisition and also responsible of the mammography analysis.
%%
The hepatic steatosis data set is analyzed by \emph{UN}, a data scientist responsible for prior analysis of the data.
%%
The third participant is \emph{TI}, a statistician with focus on epidemiology (8 years of experience) assesses the statistical reliability of the tool and the underlying methods without a focus on a specific data set.
%%
\subsection{The Hepatic Steatosis Data Set}
We employ the data set used by Niemann et al.~\cite{Niemann2014} to identify predictive features w.r.t. the reversible hepatic steatosis disorder.
%%
The dichotomous target feature is derived from the liver fat concentration measured using MRI scans.
%%
Liver fat concentrations of no more than 10\% are mapped to the `negative' class; values greater than 10\% are mapped to the `positive' class to indicate absence or presence of the disease.
%%
The data set contains labels for 578 participants. %for which the MRI data was available to that point.
%%
The MRI scans for each subject are only available in SHIP-2. %, so the target feature is only present the last study wave.

@@

As a next step, we want to introduce more regression types, which model different kinds of correlations.
%%
We also want to extend the 3D heat map to time-dependent data by expanding the difference heat map approach.
%%
We published all associated code and provide a freely accessible analysis platform open to heterogenous data types.
%%
We want to support opening up knowledge discovery to allow a diverse group of domain experts to derive insight into their data. %and support the notion of open science.
%%
\begin{small}
\acknowledgments{
%%
SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania.
%%
Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern.
%%
%The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG.
%%
This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.
%%
Sylvia Gla{\ss}er and Kai Lawonn are funded by the Federal Ministry of Education and Research within the Forschungs-campus STIMULATE under grant number '13GW0095A'.
}
\end{small}
\clearpage
\newpage
\bibliographystyle{abbrv}
%%use following if all content of bibtex file should be shown
%\nocite{*}
\bibliography{bibliography}
\end{document}
